---
title: "Assignment 3 - Part 1 - Voice In Schizophrenia"
author: "Study group 12"
date: "November 08, 2019"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 3 - Part 1 - Assessing voice in schizophrenia

Individuals with schizophrenia (SCZ) tend to present voice atypicalities. Their tone is described as "inappropriate" voice, sometimes monotone, sometimes croaky. This is important for two reasons. First, voice could constitute a direct window into cognitive, emotional and social components of the disorder, thus providing a cheap and relatively non-invasive way to support the diagnostic and assessment process (via automated analyses). Second, voice atypicalities play an important role in the social impairment experienced by individuals with SCZ, and are thought to generate negative social judgments (of unengaged, slow, unpleasant interlocutors), which can cascade in more negative and less frequent social interactions.

Several studies show *significant* differences in acoustic features by diagnosis (see meta-analysis in the readings), but we want more. We want to know whether we can diagnose a participant only from knowing the features of their voice.

The corpus you are asked to analyse is a relatively large set of voice recordings from people with schizophrenia (just after first diagnosis) and matched controls (on gender, age, education). Each participant watched several videos of triangles moving across the screen and had to describe them (so you have several recordings per person). We have already extracted the pitch once every 10 milliseconds as well as several duration related features (e.g. number of pauses, etc).

N.B. For the fun of it, I threw in data from 3 different languages: 1) Danish (study 1-4); 2) Mandarin Chinese (Study 5-6); 3) Japanese (study 7). Feel free to only use the Danish data, if you think that Mandarin and Japanese add too much complexity to your analysis.

In this assignment (A3), you will have to discuss a few important questions (given the data you have). More details below.

Part 1 - Can we find a difference in acoustic features in schizophrenia?
- Describe your sample (n of studies, n of participants, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (schizophrenia and controls) are balanced. N.B. you need to take studies into account.
- Discuss the analysis necessary to replicate the meta-analytic findings: which fixed and random effects should be included, given your dataset? E.g. what about language and study, age and gender? Discuss also how studies and languages should play a role in your analyses. E.g. should you analyze each study individually? Or each language individually? Or all together? Each of these choices makes some assumptions about how similar you expect the studies/languages to be.
- Describe the acoustic profile of a schizophrenic voice: which features are different? E.g. People with schizophrenia tend to have high-pitched voice, and present bigger swings in their prosody than controls. N.B. look also at effect sizes. How do these findings relate to the meta-analytic findings?
- Your report should look like a methods paragraph followed by a result paragraph in a typical article (think the Communication and Cognition paper)

Part 2 - Can we diagnose schizophrenia from voice only?

- Should you run the analysis on all studies and both languages at the same time? 
- Choose your best acoustic feature from part 1. How well can you diagnose schizophrenia just using it?
- Identify the best combination of acoustic features to diagnose schizophrenia using logistic regression.
- Discuss the "classification" process: which methods are you using? Which confounds should you be aware of? What are the strength and limitation of the analysis?

Bonus question: Logistic regression is only one of many classification algorithms. Try using others and compare performance. Some examples: Discriminant Function, Random Forest, Support Vector Machine, Penalized regression, etc. The packages caret and glmnet provide them. Tidymodels is a set of tidyverse style packages, which take some time to learn, but provides a great workflow for machine learning.

## Learning objectives

- Critically design, fit and report multilevel regression models in complex settings
- Critically appraise issues of replication


## Overview of part 1

In the course of this part 1 of Assignment 3 you have to:
- combine the different information from multiple files into one meaningful dataset you can use for your analysis. This involves: extracting descriptors of acoustic features from each pitch file (e.g. mean/median, standard deviation / interquartile range), and combine them with duration and demographic/clinical files
- describe and discuss your sample
- analyze the meaningful dataset to assess whether there are indeed differences in the schizophrenic voice and compare that to the meta-analysis

There are three pieces of data:

1- Demographic data (https://www.dropbox.com/s/6eyukt0r5du0xif/DemoData.txt?dl=0). It contains

- Study: a study identifier (the recordings were collected during 6 different studies with 6 different clinical practitioners in 2 different languages)
- Language: Danish, Chinese and Japanese
- Participant: a subject ID
- Diagnosis: whether the participant has schizophrenia or is a control
- Gender
- Education
- Age
- SANS: total score of negative symptoms (including lack of motivation, affect, etc). Ref: Andreasen, N. C. (1989). The Scale for the Assessment of Negative Symptoms (SANS): conceptual and theoretical foundations. The British Journal of Psychiatry, 155(S7), 49-52.
- SAPS: total score of positive symptoms (including psychoses, such as delusions and hallucinations): http://www.bli.uzh.ch/BLI/PDF/saps.pdf
- VerbalIQ: https://en.wikipedia.org/wiki/Wechsler_Adult_Intelligence_Scale
- NonVerbalIQ: https://en.wikipedia.org/wiki/Wechsler_Adult_Intelligence_Scale
- TotalIQ: https://en.wikipedia.org/wiki/Wechsler_Adult_Intelligence_Scale

2. Articulation.txt (https://www.dropbox.com/s/v86s6270w39g0rd/Articulation.txt?dl=0). It contains, per each file, measures of duration:
- soundname: the name of the recording file
- nsyll: number of syllables automatically inferred from the audio
- npause: number of pauses automatically inferred from the audio (absence of human voice longer than 200 milliseconds)
- dur (s): duration of the full recording
- phonationtime (s): duration of the recording where speech is present
- speechrate (nsyll/dur): average number of syllables per second
- articulation rate (nsyll / phonationtime): average number of syllables per spoken second
- ASD (speakingtime/nsyll): average syllable duration

3. One file per recording with the fundamental frequency of speech extracted every 10 milliseconds (excluding pauses): https://www.dropbox.com/sh/b9oc743auphzxbg/AAChUsvFc6dIQSlM9eQTL53Aa?dl=0
- time: the time at which fundamental frequency was sampled
- f0: a measure of fundamental frequency, in Herz

NB. the filenames indicate:
- Study: the study, 1-6 (1-4 in Danish, 5-6 in Mandarin Chinese)
- D: the diagnosis, 0 is control, 1 is schizophrenia
- S: the subject ID (NB. some controls and schizophrenia are matched, so there is a 101 schizophrenic and a 101 control). Also note that study 5-6 have weird numbers and no matched participants, so feel free to add e.g. 1000 to the participant ID in those studies.
- T: the trial, that is, the recording ID for that participant, 1-10 (note that study 5-6 have more)

### Getting to the pitch data

You have oh so many pitch files. What you want is a neater dataset, with one row per recording, including a bunch of meaningful descriptors of pitch. For instance, we should include "standard" descriptors: mean, standard deviation, range. Additionally, we should also include less standard, but more robust ones: e.g. median, iqr, mean absoluted deviation, coefficient of variation. The latter ones are more robust to outliers and non-normal distributions.

Tip: Load one file (as a sample) and:
- write code to extract the descriptors
- write code to extract the relevant information from the file names (Participant, Diagnosis, Trial, Study)
Only then (when everything works) turn the code into a function and use map_df() to apply it to all the files.
See placeholder code here for help.

```{r}
library(pacman)

p_load(tidyverse, lme4, groupdata2, MuMIn, Metrics) #load packages


#define function to read filenames and extract pitch data
My_f <- function(x){
                        data.frame(Study = regmatches(x, gregexpr('Study[0-9]{1}', x))[[1]], #study
                                  Diagnosis = regmatches(x, gregexpr('D[0-9]{1}', x))[[1]], #Diagnosis
                                Subject = regmatches(x, gregexpr('S[0-9]{1,3}', x))[[1]], #Subject
                                  Trial = regmatches(x, gregexpr('T[0-9]{1,2}', x))[[1]], #Trial
                          mean = mean(read.delim(x)[,2]), #mean
                          sd = sd(read.delim(x)[,2]), #sd
                          IQR = IQR(read.delim(x)[,2]), #inner quartile range
                          min = range(read.delim(x)[,2])[1], #min
                          max = range(read.delim(x)[,2])[2], #max
                          dur = range(read.delim(x)[,1])[2]) #duration
                                  }




#listing files
files <- list.files(path = "C:/Users/jacda/Desktop/Cognitive science/R-stuff/Class work 3/Assignment-3/Assignment-3-1/data/", pattern = "*.txt", full.names = T) 

#applying function the file list
l <- lapply(files, My_f)

#binding observations into data.frame
tbl <- l %>%  bind_rows()


write.csv(tbl, "pitchdata.csv", fileEncoding = "UTF-8")

```


### Now you need to merge demographic/clinical, duration and pitch data

```{r}
#reading demographic data

demo <- read.csv("DemographicData2.csv", encoding = "UTF-8")

#padding participants IDs with 0s to match participant IDs in pitch data
demo$Participant[demo$Study == 6] <- str_pad(demo$Participant[demo$Study == 6], width=3, side="left", pad="0")
demo$Participant[demo$Study == 7] <- str_pad(demo$Participant[demo$Study == 7], width=2, side="left", pad="0")


#reading duration data
arti <- read.csv("Articulation.txt", encoding = "UTF-8", stringsAsFactors = F)

#sorting duration data according to soundname (ID) column to match with pitch data
art_sor <- arti[order(arti[,1], decreasing = F ),]


#combining pitch data and duration data
pitch <- tbl
pitch[,11:18] <- art_sor[,1:8]  



demo$Study <- paste0("Study", demo$Study) #matching study variable in demographic data with pitch data
demo$Participant <- paste0("S", demo$Participant) #matching participant variable in demographic data with pitch data
demo$Diagnosis <- ifelse(demo$Diagnosis == "Control", "D0", "D1") #matching diagnosis variable in demographic data with pitch data
demo$ID <- paste0(demo$Study, demo$Diagnosis, demo$Participant) #creating unique ID variable in demographic data for merging


pitch$ID <- paste0(pitch$Study, pitch$Diagnosis, pitch$Subject)#creating unique ID variable in pitch data for merging
pitch$TID <- paste0(pitch$ID, pitch$Trial) #creating unique Trial variable in pitch data

```

Some of the duration data from study 7 is not easily combined however we have chosen to include it

Executing the following lines lets you expect the weird ID variables

gone <- which(!pitch$soundname %in% pitch$TID)

pitch[gone,]

```{r}

merged <- merge(demo, pitch, by ="ID") #merging data by the unique ID variable 
#8 observations disappear

#these are the 8 observations
pitch[which(!pitch$ID %in% demo$ID),]


#writing the final data frame
write.csv(merged_data, "merged_data.csv", fileEncoding = "UTF-8")
```


## Now we need to describe our sample

First look at the missing data: we should exclude all recordings for which we do not have complete data.
Then count the participants and recordinsgs by diagnosis, report their gender, age and symptom severity (SANS, SAPS and Social)
Finally, do the same by diagnosis and study, to assess systematic differences in studies.
I like to use group_by() %>% summarize() for quick summaries

```{r}

#Participants by diagnosis
merged %>% group_by(Diagnosis.x) %>% summarise(length(unique(ID)))


#Summarised by Gender
merged %>% group_by(Gender,Diagnosis.x) %>%
    summarise(Participants=length(unique(ID)), #participants
              Mean_Age=mean(Age,na.rm=TRUE), #age
              Mean_VIQ=mean(VerbalIQ,na.rm=TRUE), #mean verbal IQ
              SD_VIQ=sd(VerbalIQ,na.rm=TRUE), #sd verbal IQ
              Mean_SANS=mean(as.numeric(SANS), na.rm=TRUE), #mean SANS
              SD_SANS=sd(as.numeric(SANS),na.rm=TRUE), #sd  SANS
              Mean_Saps=mean(as.numeric(SAPS),na.rm=TRUE), #mean SAPS
              SD_SAPS=sd(as.numeric(SAPS),na.rm=TRUE)) #sd SAPS

#Summarised by study
merged %>% group_by(Study.x,Diagnosis.x) %>%
    summarise(Participants=length(unique(ID)), #participants
              Mean_Age=mean(Age,na.rm=TRUE), #age
              Mean_VIQ=mean(VerbalIQ,na.rm=TRUE), #mean verbal IQ
              SD_VIQ=sd(VerbalIQ,na.rm=TRUE), #sd verbal IQ
              Mean_SANS=mean(as.numeric(SANS), na.rm=TRUE), #mean SANS
              SD_SANS=sd(as.numeric(SANS),na.rm=TRUE), #sd  SANS
              Mean_Saps=mean(as.numeric(SAPS),na.rm=TRUE), #mean SAPS
              SD_SAPS=sd(as.numeric(SAPS),na.rm=TRUE)) #sd SAPS
```

## Now we can analyze the data

If we look at the meta-analysis, there are differences (measured as Hedges' g, very close to Cohen's d, that is, in standard deviations) in
- pitch variability (lower, Hedges' g: -0.55, 95% CIs: -1.06, 0.09)
- proportion of spoken time (lower, Hedges' g: -1.26, 95% CIs: -2.26, 0.25)
- speech rate (slower, Hedges' g: -0.75, 95% CIs: -1.51, 0.04)
- pause duration (longer, Hedges' g: 1.89, 95% CIs: 0.72, 3.21). (Duration - Spoken Duration) / PauseN

We need therefore to set up 4 models to replicate the findings. Feel free of course to test more features.

N.B. the meta-analytic findings are on scaled measures. If you want to compare your results with them, you need to scale your measures as well: subtract the mean, and divide by the standard deviation.
N.N.B. We want to think carefully about fixed and random effects in our model. In particular: how should study be included? Does it make sense to have all studies put together? Does it make sense to analyze both languages together? Relatedly: does it make sense to scale all data from all studies together?
N.N.N.B. If you want to estimate the studies separately, you can try this syntax:  Feature ~ 0 + Study + Study:Diagnosis + [your randomEffects]. Now you'll have an intercept per each study (the estimates for the controls) and an effect of diagnosis per each study

- Bonus points: cross-validate the models and report the betas and standard errors from all rounds to get an idea of how robust the estimates are. 

```{r}
set.seed(1)


merged <- merged_data[,c(1:14, 18:33)] #selecting variables for modelling



merged$pause_dur <- as.numeric(((merged$dur..s.- merged$phonationtime..s.)/merged$npause)) #pause duration
merged$pause_dur[merged$npause == 0] <- 0 #set observations with no pauses to 0


#Time to scale variables
#scaling variables by language as to remove variability expressed by language
merged <- merged %>%
    group_by(Language) %>%
    mutate(scaled_IQR = as.numeric(scale(IQR)), 
                  scaled_speak_time = as.numeric(scale(phonationtime..s.)),
                  scaled_speech_rate = as.numeric(scale(speechrate..nsyll.dur.)),
                  scaled_pause_dur = as.numeric(scale(pause_dur)))




#modelling IQR
model_IQR <- lmer(scaled_IQR ~ 1 + Diagnosis.x + (1|ID) + (1|Study.x), data = merged, REML = F,  control = lmerControl(
  optimizer = "nloptwrap",
  calc.derivs = F,
  optCtrl = list(
    ftol_abs = 1e-10,
    xtol_abs = 1e-10,
    maxeval = 10000
  )))

#summary model_IQR
sum_IQR <- summary(model_IQR)
sum_IQR
hedges_g(sum_IQR$coefficients[2,1], totaln = 3998) #hedges g


#k-fold cross-validate modelIQR with balanced groups

#Write function for later use
K_fold <- function(model, k_fold, data, feature){ #takes arguments for which model to use, how many folds, what data and which feature to test
  result2 <- data.frame()
for (n in 2:k_fold){
  result <- c()
  folds1 <- fold(data, n, cat_col = "Diagnosis.x", id_col = "ID")
  for (i in unique(folds1[[".folds"]])){
  test <- subset(folds1, .folds == i)
  train <- subset(folds1, .folds != i)
  mod <- model
  result[i] <- rmse(test[,feature == colnames(test)][[1]], predict(mod,test, allow.new.levels = T))}
  result2[n,1] <- mean(result)
  colnames(result2)[1] <-"mean_RMSE" 
  result2[n,2] <- n
  colnames(result2)[2] <-"k_folds" 
  result2[n,3] <- feature
  colnames(result2)[3] <-"feature" 
}
  result2
}



#Using function
res_IQR <- K_fold(model_IQR, 163, merged, "scaled_IQR")



#Displaying the lowest RMSE and the corresponding number of folds
head(res_IQR[order(res_IQR[,1], decreasing = F ),])



#modelling speak time
model_speak_time <- lmer(scaled_speak_time ~ 1 + Diagnosis.x + (1|ID) + (1|Study.x), data = merged, REML = F,  control = lmerControl(
  optimizer = "nloptwrap",
  calc.derivs = F,
  optCtrl = list(
    ftol_abs = 1e-10,
    xtol_abs = 1e-10,
    maxeval = 10000
  )))

#summary model_speak_time
sum_speak_time<- summary(model_speak_time)
sum_speak_time
hedges_g(sum_speak_time$coefficients[2,1], totaln = 3998) #hedges g

#k-fold cross-validate model speak time with balanced groups
res_speak_time <- K_fold(model_speak_time, 163, merged, "scaled_speak_time")


#Displaying the lowest RMSE and the corresponding number of folds
head(res_speak_time[order(res_speak_time[,1], decreasing = F ),])



#modelling speech rate
model_speech_rate <- lmer(scaled_speech_rate ~ 1 + Diagnosis.x + (1|ID) + (1|Study.x), data = merged, REML = F,  control = lmerControl(
  optimizer = "nloptwrap",
  calc.derivs = F,
  optCtrl = list(
    ftol_abs = 1e-10,
    xtol_abs = 1e-10,
    maxeval = 10000
  )))

#summary model_speech_rate
sum_speech_rate <- summary(model_speech_rate)
sum_speech_rate
hedges_g(sum_speech_rate$coefficients[2,1], totaln = 3998) #hedges g


#k-fold cross-validate model speech rate with balanced groups
res_speech_rate <- K_fold(model_speech_rate, 163, merged, "scaled_speech_rate")


#Displaying the lowest RMSE and the corresponding number of folds
head(res_speech_rate[order(res_speech_rate[,1], decreasing = F ),])

#modelling pause duration
model_pause_dur <- lmer(scaled_pause_dur ~ 1 + Diagnosis.x + (1|ID) + (1|Study.x), data = merged, REML = F,  control = lmerControl(
  optimizer = "nloptwrap",
  calc.derivs = F,
  optCtrl = list(
    ftol_abs = 1e-10,
    xtol_abs = 1e-10,
    maxeval = 10000
  )))

#summary model_pause_duration
sum_pause_dur <- summary(model_pause_dur)
sum_pause_dur
hedges_g(sum_pause_dur$coefficients[2,1], totaln = 3998) #hedges g

#k-fold cross-validate model pause duration with balanced groups

#Exluding missing data
merged2 <- merged[complete.cases(merged$scaled_pause_dur),]

res_IQR <- K_fold(model_pause_dur, 163, merged2, "scaled_pause_dur")


#Displaying the lowest RMSE and the corresponding number of folds
head(res_pause_dur[order(res_pause_dur[,1], decreasing = F ),])



#Combining all result data frames into a single data frame
res_frame<- rbind(res_IQR, res_speech_rate, res_speak_time, res_pause_dur)

#Plotting each models root mean squared error as a function of number folds used in cross-validation
ggplot(res_frame, aes(result2, fold, colour = feature))+
  geom_point()+
  ggtitle("K-fold crossvalidaton of scaled features with balanced groups")+
  xlab("mean RMSE")+
  ylab("Number of folds")



#Writing data to be used in part 2
write.csv(merged, "schizo_data.csv")
```

```{r}

#Model selection based on AICc using dredge


#Random effects model
#na action needs to be set to omit otherwise model will not fit
options(na.action = "na.omit")

#Defining global random effects model
glob_rf_mod <- glmer(Diagnosis.x ~ 1 + scaled_IQR + scaled_pause_dur + scaled_speak_time + scaled_speech_rate + (1|ID) + (1|Study.x), data = merged, family = binomial, REML = F,  control = glmerControl(
  optimizer = "nloptwrap",
  calc.derivs = F,
  optCtrl = list(
    ftol_abs = 1e-10,
    xtol_abs = 1e-10,
    maxeval = 10000
  )))

#Needs to be set to fail otherwise dredge won't work
options(na.action = "na.fail")

best_rf_model <- MuMIn::dredge(glob_rf_mod)

#Data frame with all possible model combination and their estimates
best_rf_model




#Model without random effetcs
#na action needs to be set to omit otherwise model will not fit
options(na.action = "na.omit")

#Defining global model
glob_mod <- glm(Diagnosis.x ~ 1 + scaled_IQR + scaled_pause_dur + scaled_speak_time + scaled_speech_rate, data = merged, family = binomial)

options(na.action = "na.fail")

best_model <- MuMIn::dredge(glob_mod)

#Data frame with all possible model combination and their estimates
best_model

```

## N.B. Remember to save the acoustic features of voice in a separate file, so to be able to load them next time


## Reminder of the report to write 

Part 1 - Can we find a difference in acoustic features in schizophrenia?
- Describe your sample (n of studies, n of participants, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (schizophrenia and controls) are balanced. N.B. you need to take studies into account.
- Discuss the analysis necessary to replicate the meta-analytic findings: which fixed and random effects should be included, given your dataset? E.g. what about language and study, age and gender? Discuss also how studies and languages should play a role in your analyses. E.g. should you analyze each study individually? Or each language individually? Or all together? Each of these choices makes some assumptions about how similar you expect the studies/languages to be.
- Describe the acoustic profile of a schizophrenic voice: which features are different? E.g. People with schizophrenia tend to have high-pitched voice, and present bigger swings in their prosody than controls. N.B. look also at effect sizes. How do these findings relate to the meta-analytic findings?
- Your report should look like a methods paragraph followed by a result paragraph in a typical article (think the Communication and Cognition paper)


